Individual and Differential Harm in Redistricting

### 2025-08-25 paper/harm.qmd
add EI sentence, 5k words

As defined above, $h_i(v,q,\tilde q)$ captures harm in a single election and relative to one counterfactual districting plan $\tilde q$.
Researchers are generally interested in outcomes over multiple elections and a much larger set of counterfactual plans.
<!-- Ideally, we would average individual harm over a set of future elections under the enacted districting plan and over a set of hypothetical districting plans that are consistent with districting criteria.
We could also consider averaging over a set of individuals with some shared characteristics. -->

@@

The randomness in $V$ and $\tilde Q$ allows us to compare the enacted districting plan to many plausible counterfactual alternatives, and across a range of electoral environments.
At times we may wish to study on a particular election or a particular counterfactual plan.
Treating $V$ and $\tilde Q$ as random variables allows us to condition on particular values of each.
A particular specification of the distribution of $V$ and $\tilde Q$ is required only when *estimating* our harm measures in a particular application, a topic we return to in the methods section.

@@

To that end, let $\Groups=\{G_1,G_2,\dots\}$ be a (random) partition of the set of voters.
In general we will not require the group identities to be fixed with respect to elections.[^1]
This is because some group identities, like partisan preference, may themselves change election-to-election.

@@

Zero differential harm between groups is analogous to the notion of *error rate balance* in the algorithmic fairness literature, which seeks to equalize false positive and negative rates across groups [@zafar2017fairness; @chouldechova2017fair].
There, the implicit counterfactual is a perfect predictor.
Compared to the perfect predictor, the actual process harms individuals whenever it makes a false positive or negative error.
Here, there can be no perfect districting plan, and the counterfactual instead consists of a collection of plausible alternatives.

>>>>

As defined above, $h_i(v,q,\tilde q)$ captures harm in a single election and relative to one counterfactual districting plan $\tilde q$.
Researchers are generally interested in outcomes over multiple elections and a larger set of counterfactual plans.
<!-- Ideally, we would average individual harm over a set of future elections under the enacted districting plan and over a set of hypothetical districting plans that are consistent with districting criteria.
We could also consider averaging over a set of individuals with some shared characteristics. -->

@@

The randomness in $V$ and $\tilde Q$ allows us to compare the enacted districting plan to many plausible counterfactual alternatives, and across a range of electoral environments.
We may also wish to study a particular election or counterfactual plan.
Treating $V$ and $\tilde Q$ as random variables allows us to condition on particular values of each.
A particular specification of the distribution of $V$ and $\tilde Q$ is required only when *estimating* our harm measures in a particular application, a topic we return to in the methods section.

@@

To that end, let $\Groups=\{G_1,G_2,\dots\}$ be a (random) partition of the set of voters.
In general we do require the group identities to be fixed with respect to elections.[^1]
This is because some group identities, like partisan preference, may themselves change election-to-election.
Groups may need to be estimated with methods such as ecological inference [@king1997solution], as discussed in @sec-estimating.

@@

Zero differential harm between groups is analogous to the notion of *error rate balance* in the algorithmic fairness literature, which seeks to equalize false positive and negative rates across groups [@zafar2017fairness; @chouldechova2017fair].
There, the implicit counterfactual is a perfect predictor and the actual process harms individuals whenever it makes an incorrect prediction.
Here, there can be no perfect districting plan, and the counterfactual instead consists of a collection of plausible alternatives.
